---
title: "Download NEON Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{download-neon-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = "~/zhulab/neonSoilMicrobeProcessing/" # Update as necessary. Should refer to the absolute filepath of the project root directory (e.g. .../neonSoilMicrobeProcessing)
)
```


This vignette demonstrates how to use the functions and parameters in this package to download the specific scope NEON soil microbe marker gene sequence data and associated data relevant to your analysis. 

# Dependencies

To begin, ensure that you have installed all dependencies.

From CRAN:
```{r, eval=FALSE}
install.packages("dplyr")
install.packages("neonUtilities")
install.packages("R.utils")
```

```{r, message=FALSE}
library(dplyr)
library(neonUtilities)
library(R.utils)
```

In addition, source files associated with this package:

```{r}
source("./R/utils.R")
source("./code/params.R")
```

# Set up directories

Set up the directory structure associated with the various NEON data products. The structure will depend on what you have written in `params.R`. For example:

```
PRESET_OUTDIR = "/data/NEON"
PRESET_OUTDIR_SEQUENCE = "raw_sequence"
PRESET_OUTDIR_SEQMETA = "sequence_metadata"
PRESET_OUTDIR_SOIL = "soil"
```

This will generate (recursively) the following output directories:

```
/data/NEON/raw_sequence
/data/NEON/sequence_metadata
/data/NEON/soil
```

```{r}
makeOutputDirectories()
```

# Download data

## Download metadata

We begin by downloading the metadata associated with the collection and analysis of NEON soil microbial samples, using the `downloadSequenceMetadata()` function.

The `downloadSequenceMetadata()` function downloads NEON data product DP1.10108 ("Soil microbe marker gene sequences") and assembles it using the `neonUtilities` package. This data product contains a number of tables related to accessing raw sequence data. A joined table containing information about the raw data URLs and the marker gene sequencing parameters is returned by this function. 

If time and space limitations are not issues for you, you can download the entire dataset over the course of a few hours. However, `downloadSequenceMetadata()` accepts arguments that can be used to specify the subset of the data you are interested in keeping for your analysis.

In the following example, we run `downloadSequenceMetadata()` with a number of arguments to narrow the range of data to be downloaded.

```{r}
meta <- downloadSequenceMetadata(startYrMo = "2018-03", endYrMo = "2018-07", 
                                 sites = c("CLBJ", "CPER", "OSBS"),
                                 targetGene = "ITS")
```

## Download raw sequence data

Now that we have the metadata table loaded into memory, we retrieve a table of unique raw data files and their sequencing run IDs. 

```{r, message=1:5}
download_success <- downloadRawSequenceData(meta)
```

## Reorganize and rename the downloaded files

Get the names of all downloaded files.

```{r}
outdir_sequence <- file.path(PRESET_OUTDIR, PRESET_OUTDIR_SEQUENCE)
fn0 <- file.path("./NEON/raw_sequence", meta$rawDataFileName)
fn <- unique(fn0[file.exists(fn0)])
```

The following function untars each file (if necessary), appends the sequencer run ID to the beginning of each filename, and moves files to subfolders within your  sequence directory according to their target gene. The terminal directory name for either target gene will be `0_raw`.

```{r}
reorganized_files <- organizeRawSequenceData(fn, meta)
head(reorganized_files)
```

